\customchapter{ABSTRACT}

Air quality in enclosed spaces, especially in university classrooms, is crucial for the well-being and performance of students. High levels of carbon dioxide (CO\(_2\)) and particulate matter (dust), along with variations in temperature and humidity, can negatively affect the health and cognitive performance of occupants. In this context, this work presents the design and implementation of a multivariable real-time environmental monitoring and alert system based on an Internet of Things (IoT) network architecture. The system is composed of sensor nodes that collect CO\(_2\), dust, temperature, and humidity measurements, transmitting them via LoRa wireless communication to a custom gateway based on the ESP32 microcontroller. This gateway acts as a bridge to send the data to a cloud infrastructure, deployed on Microsoft Azure using Infrastructure as Code (IaC). The web platform for data visualization was developed using React, coupled with an API hosted on Azure Container Apps, and the collected information is stored in an Azure SQL Database. The system includes automated alerts and an active mitigation subsystem that compares the performance of various ventilation control strategies (PID, Fuzzy, LQR) when permissible thresholds are exceeded, contributing to the creation of safer educational environments. Integration tests validated the complete data flow and its ability to monitor and actuate mitigation measures in real-time.

\noindent \textbf{Keywords:}\\
\noindent Air quality; CO\(_2\); particulate matter; temperature; humidity; IoT; LoRa; Azure; comparative control; automated ventilation.
